Thermal reactionomes reveal adaptive responses to thermal extremes in warm

### 2014-10-09 doc/ApTranscriptome_manuscript.Rmd
edits through end of Results

Thermal tolerance is a key trait governing the distributions of species and is likely useful in predicting their future responses to climate change. To understand how plasticity of gene expression underlies thermal tolerance, we describe the *thermal reactionome*, the reaction norm for all genes in an organism's transcriptome measured across an experimental thermal gradient. We characterize the reactionomes of two common ant species in the eastern U.S: the northern cold-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*. We found that at least 10% of genes were thermally responsive. The warm-adapted *A. carolinensis* displayed higher mean levels of expression for responsive genes, exhibited greater inertia of gene expression change and upregulated fewer heat shock proteins than did *A. picea* in response to thermal extremes. This pattern is consistent with hypotheses of constitutive thermal defense and enhanced proteome stability. *Aphaenogaster carolinensis* also down-regulated expression of more genes in response to thermal extremes than did *A. picea*. A subset of these same genes were upregulated in *A. picea*, including genes mapped to ontology terms such as metabolism, cellular organization, glycoprotein synthesis and DNA repair. These differences indicate that the two closely related species employ different strategies for coping with temperature stress. The northern *A. picea* actively resists thermal shock by synthesizing heat shock proteins and other protective molecules to maintain function, whereas the southern *A. carolinensis* slows down metabolism and relies upon other mechanisms to tolerate thermal stress. These results are in line with expectations that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

@@

Coping with progressive changes in climate requires both short-term physiological acclimation and long-term adaptive changes. To better understand the relationship between these two timescales, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A. carolinensis*. Surprisingly, we found divergent transcriptomic responses in these species. *Aphaenogaster picea* responded to high and low temperatures through upregulation of a larger proportion of its transcriptome with a robust heat shock response. In contrast, *A. carolinensis* required more extreme temperatures to trigger a protective response and downregulated genes related to metabolism. These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely encountered stressors, which may limit adaptive potential of northern species in the face of rapidly shifting climatices.

@@

To address this gap of knowlege, we need information on the extent to which selection has acted upon the diversity genes and their plasticity of expression involved in thermal tolerance [@hoffmann2008; @somero2011]. In recent years, whole organism gene expression approaches (e.g. transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes involved in thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014] and other traits. Such studies typically use an ANOVA-type experimental or sampling design, with only a few environmental levels, and often find only a few dozen to hundred genes with differential expression in different thermal regimes. However, temperature and other environmental factors vary continuously in nature. As a result, such comparisons (e.g. hot vs. cold) are likely to miss key differences that are due not just to whether it is hot, but rather how hot it is. Continuous variation is better characterized with a reaction norm approach, which describes variation in the phenotype of a single genotype across an environmental gradient [@gomulkiewicz1992]. Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values at both the species and population levels [@murren2014].

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* experiences significantly warmer and less seasonally variable temperatures than *A. picea* (...data...), with correspondingly higher minimal and maximal critical thermal limits (...CTmin and CTmax from HF/DF...). Experimental warming experiments indicate that *Aphaenogaster* abundance and competitive ability may be significantly impacted by temperature increases (...refs...).

To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Material). We used these data to quantitatively test a number of alternative mechanistic hypotheses to explain how populations evolutionarily increase their thermal breadth in response to stressful environments. First, the *enhanced defense* hypothesis (...ref...) proposes that the stress-adapted species recruits additional molecular pathways to provide greater protection from damage. This hypothesis predicts that *A. picea* induces more genes in response to cold temperatures, whereas *A. carolinensis* induces more genes in response to high temperatures. An additional prediction of this hypothesis is that upregulation of the shared molecular pathways should be strong or occur in response to less initial temperatures stress for more commonly-experienced thermal extremes; low temperatures for *A. picea*, high temperatures for *A. carolinensis*. Second, the *genetic assimiliation* hypothesis [@waddington1953; @sikkink2014] proposes that exposure to more frequent stressors selects for a shift from inducible to constitutive expression of stress-response genes. This hypothesis predicts that transcripts responsive to high temperatures in *A. picea* will have higher constitutive expression in *A. carolinensis*, while transcripts responsive to low temperatures in *A. carolinensis* will have higher constitutive expression in *A. picea*. Finally, the *tolerance* hypothesis (...ref...fields 2001?...) proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural [@fields2001] or behavioral [@gordon2013; @sunday2014] adapations. This hypothesis predicts that the two species should differ in the identity and timing of gene induction in response to temperature changes. Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adapatation as the proportion of the transcriptome matching each of these predicted patterns.

@@

## Thermally responsive gene responsed non-linearly to changes in temperature

The combined *Aphaenogaster* transcriptome assembly contained 98,186 transcripts. About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995]. Of these, `r prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach. The result is the same if we focus only on those transcript with a BLAST hit (4,618 transcripts out of 51,246). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate is a lower-bound for the true proportion of genes that are thermally responsive.

@@

## *A. picea* recruits additional molecular pathways in response to heat stress

Transcripts varied both qualitatively and quantitatively in their pattern of response to the temperature gradient. We grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C). The expression pattern differed between species for `r prettyNum(sum(sigtable[sigtable$Coefficient %in% "Temperature:Colony", "Number.significant"]), big.mark = ",", scientific = FALSE)` (`r prettyNum(sigtable[sigtable$Coefficient == "Temperature:Colony", "Number.significant"] / sigtable[sigtable$Coefficient == "Total", "Number.significant"]*100, digits=2)`%) of the thermally-responsive transcripts.

Consistent with the *enhanced response* hypothesis, *A. picea* induced expression of about 1,000 (30%) more transcripts than *A. carolinensis* in response to cold temperatures (Table 1, ...stats...). However, *A. picea* also induced expression of more transcripts, though to a smaller extent (~14%), than did *A. carolinensis* in response to high temperatures and induced more transcripts that are responsive to both high low temperatures (*Bimodal* transcripts; Table 1).

The additional genes recruited included molecular processes such as...GO terms...

## *A. carolinensis* shuts-down gene expression in response to thermal extremes

While *A. picea* induced more genes, the total number of thermally-responsive transcripts did not differ between species (...stat...). Instead, most of these transcripts shifted from one responsive category to another. These shifts were not randomly distributed (...Stuart-Maxwell test of marginal homogeneity ...) indicating that these differences are not simply due to neutral processes. **Intermediate** transcripts in *A. carolinensis* were significantly more likely to have **Low** or **Bimodal** expression in *A. picea* than expected by chance (...stat...), while there was a significant deficit of the reverse shift of *A. picea* **Intermediate** transcripts to any upregulated category in *A. carolinensis* (Fig. 1). Similarly, both **High** and **Bimodal** transcripts in *A. picea* were more likely to be **Not Responsive** in *A. carolinensis* than expected, while the reverse was not significant (Fig. 1).

The molecular processes in the set of transcripts that switched from down-regulation in *A. carolinensis* to up-regulation in *A. picea* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` GO terms, of which ... were related to metabolism. In addition, there were GO terms related to repair (e.g. "histone modification"), transcript regulation (e.g. "positive regulation of cellular process") and cellular organization (e.g. "transport", "organelle organization"). Altogether, these results suggest that in response to thermal extremes, the warmer-climate species, *A. carolinensis*, primarily shuts down metabolism while the colder-climate species, *A. picea*, increases metabolism and initiates damage repair responses.

@@

## Greater constitutive expression of thermally-responsive genes in *A. carolinensis*

We tested the *genetic assimilation* hypothesis, a shift from inducible to constitutive expression of stress-response genes, by comparing constitutive expression, here defined as expression at 19.5°C, between species for high temperature responsive transcripts in *A. picea* and *A. carolinensis*. Consistent with this hypothesis, we found greater constitutive expression in the warm-climate species *A.carolinensis* than *A. picea* for high temperature responsive transcripts in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`, 95% CI location shift `r paste(round(w1$conf.int[1], 3), "--", round(w1$conf.int[2], 3))`). However, this pattern was also true for low temperature responsive transcripts in *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`, 95% CI location shift `r paste(round(w2$conf.int[1], 3), "--", round(w2$conf.int[2], 3))`), and **Intermediate** transcripts in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`, 95% CI location shift `r paste(round(w3$conf.int[1], 3), "--", round(w3$conf.int[2], 3))`). Thus, the overall pattern is that thermally-responsive transcripts in *A. carolinensis* had higher constitutive expression than those in *A. picea*.

An exception to this overall pattern is the set of genes that had **High** or **Bimodal** expression in *A. picea* but were not thermally responsive in *A. carolinensis* (blue cells in bottom row of Fig. 1). Rather than having constitutively elevated expression as predicted, these genes were not expressed (...mean TPM...) in *A. carolinensis*.

@@

To test the *tolerance* hypothesis, we examined the timing of gene induction in response to temperature changes, and the identity of these genes. Specifically, we compared the mean temperatures of transcript activation, defined as the temperature at which the transcript showed the greatest positive change in expression, between species. **High** and **Low** transcripts were included only for the direction in which they were upregulated, while **Bimodal** genes wre included for analysis at both high and low temperature activation. In support of the *tolerance* but not the *enhanced defense* hypothesis, transcripts turned on at high temperatures in *A. carolinensis* were activated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A. picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For low temperatures, there was no difference in the mean temperature of transcript activation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~`r round(T_on.low.ttest$parameter, 0)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2).

Similarly, for the **Intermediate** transcripts we compared the mean temperatures of down-regulation of transcript expression between species at both high (> 21°) and low (< 21°C) temperatures. Overall, the thresholds for down-regulation were shifted toward cooler temperatures than the corresponding thresholds for gene activation, with a shift of approximately ... at the higher temperature extremes and ...  at low temperature extremes. Again, consistent with the *tolerance* hypothesis, we found greater inertia of gene expression to changes in temperature for *A. carolinensis* than *A. picea* drive by a differential response to heat stress. With increasing temperatures, **Intermediate** transcripts were down-regulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~df=`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), while with decreasing temperatures the species did not differ in mean temperature of down-regulation (*A. carolinensis* `r round(T_off.int.low.ttest$estimate[1], 1)` compared to *A. picea* `r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~df=`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).

@@

Under the *enhanced defense* hypothesis, we predicted that the cool-climate species *A. picea* would induce more genes in response to low temperatures than *A. carolinensis*, while *A. carolinensis* would induce more genes in response to high temperatures. This prediction was only true for *A. picea*, which induced the same genes and then recruited additional genes in response to both low and high temperatures (Fig. 1). In addition, we predicted that upregulation should occur in response to less extreme cold and high temperatures for *A. picea* and *A. carolinensis*, respectively. In contrast, we found no difference between species in the mean temperature of expression response for cold temperatures (Fig. 2A,C), and that higher temperatures were required to induce the expression response for *A. carolinensis* than *A. picea* for cold temperatures. While conflicting with the *enhanced defense* hypothesis, this is consistent with the *tolerance* hypothesis for *A. carolinensis*.

@@

> We examined the reaction norms for transcripts in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig. 3). Although `r resp.stress.responsive.count` transcripts were annotated with this GO term among the thermally responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r round(GO.stress.Xsq$statistic, 2)`, P < `r round(GO.stress.Xsq$p.value, 2)`). In addition, of the `r nrow(hsp_all)` genes annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to cold or hot temperatures in *A. picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*.

Given the differences in the patterns thermal-responsiveness between species (Fig. 2), it is worth noting the similarity in expression patterns between species. In both species, there are 3-4 $\times$ more transcripts upregulated at cold than high temperatures  (Fig. 3). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures. At least among *Drosophila* taxa, limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012]. In this light, for *A. picea* and *A. carolinensis*, our results suggest that adaptability to increased temperatures is more limited than to cold temperatures, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].

@@

In this work, we have brought reaction norms to the genomic era by characterizing the thermal reactionomes of two temperate ant species, *Aphaenogaster picea* and *A. carolinensis*. At least 10% of their transcriptomes are thermally responsive. For both species, more genes responded to cold than hot temperatures. Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have less adaptive potential in response to hot than cold temperatures. Our results also indicate that these two ant species have different responses to thermal extremes. *A. picea* responds by increasing metabolism and activating heat shock proteins and other protective molecules, whereas *A. carolinensis* shuts down metabolism and must rely on other mechanisms for thermal tolerance. The thermal reactionomes of these two species provide key insights into the genetic basis of thermal tolerance, and a resource for the future study of ecological adaptation in ant species. Finally, the reactionome itself illustrates a new direction for characterizing acclimation and adaptation in a changing climate.

>>>>

**Limit 250 words, at 250! If you add words...find others to delete!**

Thermal tolerance is a key trait governing the distributions of species and is useful in predicting their responses to climate change. To quantitatively evaluate among three hypotheses for the evolution of thermal tolerance, we describe a novel analysis of the *reactionome*; the reaction norm for all genes in an organism's transcriptome measured across an experimental gradient. We characterized the thermal reactionomes of two common ant species in the eastern U.S: the northern cool-climate *Aphaenogaster picea* and the southern warm-climate *Aphaenogaster carolinensis*. We found that at least 8% of genes were thermally responsive. The cool-adapted *A. picea* induced expression of more genes in response to extreme temperatures than *A. carolinensis*, including heat shock proteins, consistent with the *enhanced defense* hypothesis. In contrast, *A. carolinensis* displayed higher constitutive expression of responsive genes and required greater changes towards high, but not low, temperatures to induce changes in gene expression than did *A. picea*. This pattern was consistent with *tolerance* hypothesis at high, but not low, temperatures. We found no evidence for the *genetic assimilation* hypothesis as there was no trade-off between constitutive and inducible levels of gene expression. Overall, our results indicate that the northern *A. picea* actively resists thermal shock by synthesizing heat shock and other protective molecules to maintain function, whereas the southern *A. carolinensis* relies upon other mechanisms to tolerate thermal stress. These results are in line with expectations that thermally stressful habitats select for an evolutionary shift in response mechanism away from high-cost damage repair and toward lower-cost prevention.

@@

**Limit 120 words, at 120! If you add words...find others to delete!**

Coping with progressive changes in climate requires both short-term physiological acclimation and long-term adaptive changes. To better understand the relationship between these two timescales, we compared gene expression as a function of temperature in two closely-related ant species: the cool-climate *Aphaenogaster picea* and the warm-climate *A. carolinensis*. Surprisingly, we found divergent transcriptomic responses in these species. *Aphaenogaster picea* responded to high and low temperatures through upregulation of a larger proportion of its transcriptome with a robust heat shock response. In contrast, *A. carolinensis* required more extreme temperatures to trigger a protective response and downregulated genes related to metabolism. These results suggest that adaptation to chronic high temperatures involves qualitatively different mechanisms than those used to cope with rarely encountered stressors.

@@

To address this gap of knowlege, we need information on the extent to which selection has acted upon the diversity genes and their plasticity of expression involved in thermal tolerance [@hoffmann2008; @somero2011]. In recent years, whole organism gene expression approaches (e.g. transcriptomics) using high-throughput RNA sequencing (RNAseq) technology have been widely applied to identify genes involved in thermal tolerance [@meyer2011; @teets2012; @barshis2013; @oneil2014] and other traits. Such studies typically use an ANOVA-type experimental or sampling design, with only a few environmental levels, and often find only a few dozen to hundred genes with differential expression in different thermal regimes. However, temperature and other environmental factors vary continuously in nature. As a result, such comparisons (e.g. high vs. low temperatures) are likely to miss key differences that are due not just to whether it is hot, but rather how hot it is. Continuous variation is better characterized with a reaction norm approach, which describes variation in the phenotype of a single genotype across an environmental gradient [@gomulkiewicz1992]. Reaction norms differ not only in mean values, but also in their shapes [@hoffmann2013; @murren2014], and differences in the shape of reaction norms are often greater than differences in mean values at both the species and population levels [@murren2014].

Here, we extend the reaction norm approach to RNAseq analysis with the *reactionome*, which we define as a characterization of the reaction norm for all genes in an organism's transcriptome across an environmental gradient. We use this method to identify genes that are thermally responsive in two closely related eastern North American ant species, *Aphaenogaster carolinensis* and *A. picea* [@umphrey1996]. *Aphaenogaster* are some of the most common ant in eastern North America [@king2013] where they are keystone seed dispersers [@ness2009; @zelikova2011; @rodriguez-cabal2012]. *Aphaenogaster carolinensis* (mean annual temperature at collection site = ...) experiences significantly warmer and less seasonally variable temperatures than *A. picea* (mean annual temperature from collection site = ; [@hijmans2005]), with correspondingly higher minimal and maximal critical thermal limits that allow us to investigate adaptation to both lower and upper thermal extremes (...CTmin and CTmax from HF/DF...) [@diamond2012].

To characterize the thermal reactionome, we collected whole organism RNAseq data at 12 temperatures (every 3.5°C between 0 and 38.5°C) using Illumina short read sequencing technology, generated the *de novo* transcriptome using Trinity [@grabherr2011], quantified gene expression using Sailfish [@patro2014], and characterized the reaction norm for each gene using a custom analysis (Supplemental Material). We used these data to quantitatively test three mechanistic hypotheses of thermal adaptation. First, the *enhanced defense* hypothesis (...ref...) proposes that the stress-adapted species extend their thermal limits through a stronger induced response to provide greater protection from damage. This hypothesis would predict that *A. picea* induces more genes in response to low temperatures, whereas *A. carolinensis* induces more genes in response to high temperatures. Additionally, upregulation of molecular pathways shared by both species should be stronger or occur in response to less initial temperatures stress for more commonly-experienced thermal extremes; low temperatures for *A. picea*, high temperatures for *A. carolinensis*. Second, the *genetic assimiliation* hypothesis [@waddington1953; @sikkink2014] proposes that exposure to more frequent stressors selects for a shift from inducible to constitutive expression of stress-response genes. This hypothesis predicts that transcripts responsive to high temperatures in *A. picea* will have higher constitutive expression in *A. carolinensis*, while transcripts responsive to low temperatures in *A. carolinensis* will have higher constitutive expression in *A. picea*. Finally, the *tolerance* hypothesis (...ref...fields 2001?...) proposes that existing inducible stress responses become insufficient or prohibitively costly as environmental stressors increase in frequency, resulting in a shift away from induced response in favor of structural [@fields2001] or behavioral [@gordon2013; @sunday2014] adapations. This hypothesis predicts adaptation to stress should be associated with lower transcriptional responsiveness and less sensitivity to temperature perturbation, as well as a shift to an alternate suite of tolerance genes and pathways. Of course, these hypotheses are not mutually exclusive. By taking a reactionome approach, we can quantify the relative importance of each of these mechanisms to thermal adapatation as the proportion of the transcriptome matching each of these predicted patterns.

@@

## Reaction norms of thermally-responsive transcripts differ between species

The combined *Aphaenogaster* transcriptome assembly contained 98,186 transcripts. About half of these (51,246) transcripts had a signficant BLAST hit, of which 50% (25,797) had a top hit to Insecta and 37% (18,854) had a top hit to Formicidae. Of these, we identified `r prettyNum(sum(sigtable[sigtable$Coefficient %in% c("Temperature", "Temperature:Colony"), "Number.significant"]), big.mark = ",", scientific = FALSE)` transcripts, or 8% of the total, with expression patterns that were thermally responsive at a 5% False Discovery Rate [@benjamini1995]. Of these, `r prettyNum(length(quadratic.lms), big.mark = ",")` had a non-linear relationship with temperature that would likely have been missed with a standard differential expression approach. The result is the same if we focus only on those transcript with a BLAST hit (4,618 transcripts out of 51,246). However, as with all *de novo* transcriptome assemblies, this assembly is fragmented due to partial contigs and alternative transcripts [@vijay2013] suggesting that this estimate is likely a lower-bound for the true proportion of genes that are thermally responsive.

@@

The total number of thermally-responsive transcripts did not differ between species (...stat...). Within this set, there was significant variation in response both among transcripts and between the two species. Transcripts varied both qualitatively and quantitatively in their their thermal reaction norms. We grouped the thermally-responsive transcripts for each species into five expression categories based on their reaction norms: transcripts that had greatest expression at high (> 31°) temperatures (**High**), low (< 10°C) temperatures (**Low**), intermediate (10 - 30°C) temperatures (**Intermediate**), both high and low temperatures (**Bimodal**) or were not thermally responsive in that species (**Not Responsive**). For the **Bimodal** group, we required that expression at both low and high temperatures was at least one standard deviation greater than the expression at midpoint (19.5°C). The majority of the responsive genes were in the **Low** category for both species (42% and 54% for *A. carolinensis* and *A. picea*, respectively), followed by **Intermediate** (29%), **High** (12%),**Bimodal** (10%) and **Not Responsive** (8%) for *A. carolinensis*, and **Bimodal** (17%), **High** (14%), **Intermediate** (10%) and **Not Responsive** (6%) for *A. picea*.

## Expression response to thermal extremes differs between species

Despite their similar total numbers of responsive transcripts, the two species differed in the proportions of transcripts in each response category (Table 1). *Aphaenogaster picea* upregulated significantly more genes in response to both temperature extremes (*Bimodal*, ...stats...), while *A. carolinensis* downregulated more transcripts under these conditions (*Intermediate*, ...stats...). Consistent with the *enhanced defense* hypothesis, *A. picea* was more responsive specifically to low temperatures, inducing expression of 1,146 (30%) more transcripts than *A. carolinensis* (Table 1, ... stats...). At high temperatures, *A. carolinensis* showed the opposite pattern, matching the predictions of the *tolerance* hypothesis, inducing expression of ~14% fewer transcripts than *A. picea* (Table 1).

In addition to comparing the overall number of transcripts in each category, we also examined the extent to which reaction norms for individual transcripts were shared between species. Expression patterns differed for 6,792 (65%) of the thermally-responsive transcripts. Most of these shifted from one responsive category to another. These shifts were not randomly distributed (...Stuart-Maxwell test of marginal homogeneity ...) indicating that these differences are not simply due to neutral processes. **Intermediate** transcripts in *A. carolinensis* were significantly more likely to have **Low** or **Bimodal** expression in *A. picea* than expected by chance (...stat...), while there was a significant deficit of the reverse shift of *A. picea* **Intermediate** transcripts to any upregulated category in *A. carolinensis* (Fig. 1). Similarly, both **High** and **Bimodal** transcripts in *A. picea* were more likely to be **Not Responsive** in *A. carolinensis* than expected, while the reverse was not significant (Fig. 1).

The molecular processes in the set of transcripts that switched from down-regulation in *A. carolinensis* to up-regulation in *A. picea* included `r nrow(Ar.int.A22.low.gsea) + nrow(Ar.int.A22.bim.gsea)` GO terms, of which ... were related to metabolism. In addition, there were GO terms related to repair (e.g. "histone modification"), transcript regulation (e.g. "positive regulation of cellular process") and cellular organization (e.g. "transport", "organelle organization"). Altogether, these results suggest that in response to both thermal extremes, the warmer-climate species, *A. carolinensis*, primarily shuts down metabolism while the cooler-climate species, *A. picea*, increases metabolism and initiates damage repair responses.

@@

## No evidence for *genetic assimilation*

We tested the *genetic assimilation* hypothesis, a shift from inducible to constitutive expression of stress-response genes, by testing for each temperature extreme whether the suite of transcripts induced in response to that extreme showed a higher constitutive level of expression, here defined as expression at 19.5°C, in the species better adapted to these conditions. Consistent with this hypothesis, we found greater constitutive expression in the warm-climate species *A.carolinensis* than *A. picea* for high temperature responsive transcripts in *A. picea* (Wilcoxon V = `r prettyNum(w1$statistic)`, *P* < `r max(w1$p.value, 0.001)`). However, the reverse pattern of higher constitutive expression in *A. picea* was not observed for low temperature responsive transcripts in *A. carolinensis* (Wilcoxon V = `r prettyNum(w2$statistic)`, *P* < `r max(w2$p.value, 0.001)`).

Instead, the overall pattern is that thermally-responsive transcripts in *A. carolinensis*, regardless of expression pattern, had higher constitutive expression than those in *A. picea*, including those with **Intermediate** expression profiles in both species (Wilcoxon V = `r prettyNum(w3$statistic)`, *P* < `r max(w3$p.value, 0.001)`). An important exception to this pattern is the set of transcripts that had **High** or **Bimodal** expression in *A. picea* but were not thermally responsive in *A. carolinensis* (blue cells in bottom row of Fig. 1). Rather than having constitutively elevated expression as predicted, these genes were uniformally not expressed (...mean TPM...) in *A. carolinensis*.

@@

As an additional test of the *tolerance* hypothesis, we examined the timing of gene induction in response to temperature changes. We compared the mean temperatures of transcript upregulation, defined as the temperature at which the transcript showed the greatest positive change in expression, between species. **High** and **Low** transcripts were included only for the direction in which they were upregulated, while **Bimodal** genes were included for analysis at both high and low temperatures. In support of the *tolerance* but not the *enhanced defense* hypothesis, transcripts turned on at high temperatures in *A. carolinensis* were upregulated on average at `r round(T_on.high.ttest$estimate[1], 1)`°C compared to `r round(T_on.high.ttest$estimate[2], 1)`°C for *A. picea* (*t*~`r round(T_on.high.ttest$parameter, 0)`~ = `r round(T_on.high.ttest$statistic, 1)`, *P* < `r max(T_on.high.ttest$p.value, 0.001)`; Fig. 2). For low temperatures, there was no difference in the mean temperature of transcript upregulation (`r round(T_on.low.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_on.low.ttest$estimate[2], 1)`°C for *A. picea*; *t*~`r prettyNum(T_on.low.ttest$parameter, digits=2)`~ = `r round(T_on.low.ttest$statistic, 1)`, *P* < `r round(max(T_on.low.ttest$p.value, 0.001), 3)`; Fig. 2).

Similarly, for the **Intermediate** transcripts we compared the mean temperatures of down-regulation of transcript expression between species at both high (> 21°) and low (< 21°C) temperatures. Overall, the thresholds for down-regulation were shifted toward cooler temperatures than the corresponding thresholds for gene activation, with a shift of approximately ... at the higher temperature extremes and ...  at low temperature extremes. Again, we found greater inertia of gene expression only in response to high temperature extremes. With increasing temperatures, **Intermediate** transcripts were downregulated at `r round(T_off.int.high.ttest$estimate[1], 1)`°C for *A. carolinensis* compared to `r round(T_off.int.high.ttest$estimate[2], 1)` for *A. picea* (*t*~`r round(T_off.int.high.ttest$parameter, 0)`~ = `r round(T_off.int.high.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.high.ttest$p.value, 0.001), 3)`), while with decreasing temperatures the species did not differ in mean temperature of downregulation (*A. carolinensis* `r round(T_off.int.low.ttest$estimate[1], 1)` compared to *A. picea* `r round(T_off.int.low.ttest$estimate[1], 1)`; *t*~`r round(T_off.int.low.ttest$parameter, 0)`~ = `r round(T_off.int.low.ttest$statistic, 1)`, *P* < `r round(max(T_off.int.low.ttest$p.value, 0.001), 3)`).

@@

Under the *enhanced defense* hypothesis, we predicted that the cool-climate species *A. picea* would induce more genes in response to low temperatures than *A. carolinensis*, while *A. carolinensis* would induce more genes in response to high temperatures. This prediction was only true for *A. picea*, which induced the same genes and then recruited additional genes in response to both low and high temperatures (Fig. 1). In addition, we predicted that upregulation should occur in response to less extreme low and high temperatures for *A. picea* and *A. carolinensis*, respectively. In contrast, we found no difference between species in the mean temperature of expression response for low temperatures (Fig. 2A,C), and that higher temperatures were required to induce the expression response for *A. carolinensis* than *A. picea* for low temperatures. While conflicting with the *enhanced defense* hypothesis, this is consistent with the *tolerance* hypothesis for *A. carolinensis*.

@@

> We examined the reaction norms for transcripts in the GO category "response to stress" which we *a priori* expected to have significant thermal responses (Fig. 3). Although `r resp.stress.responsive.count` transcripts were annotated with this GO term among the thermally responsive trancripts, this number was not enriched compared to the total number in the transcriptome ($\chi^2$ = `r round(GO.stress.Xsq$statistic, 2)`, P < `r round(GO.stress.Xsq$p.value, 2)`). In addition, of the `r nrow(hsp_all)` genes annotated to heat shock proteins in the transcriptome, `r length(which(hsp_responsive$A22.type %in% c("Bimodal", "High", "Low")))` were upregulated to low or high temperatures in *A. picea* while only `r length(which(hsp_responsive$Ar.type %in% c("Bimodal", "High", "Low")))` were upregulated in *A. carolinensis*.

Given the differences in the patterns thermal-responsiveness between species (Fig. 2), it is worth noting the similarity in expression patterns between species. In both species, there are 3-4 $\times$ more transcripts upregulated at low than high temperatures  (Fig. 3). A potential explanation for this pattern is that selection has been stronger on upper than lower thermal limits, thereby reducing genetic variation at high temperatures [@kingsolver2013; @kelly2013]. This explanation is consistent with the observation in *Aphaenogaster rudis* [@warren2013] and other ectotherms [@addo-bediako2000; @hoffmann2013] that critical maximum temperatures vary less among taxa than do critical minimum temperatures. At least among *Drosophila* taxa, limits to thermal maxima appear to be due to a scarcity of genetic variation in thermal tolerance [@kellermann2012]. In this light, for *A. picea* and *A. carolinensis*, our results suggest that adaptability to increased temperatures is more limited than to low temperatures, and these species risk replacement by more heat-tolerant competitors [@rodriguez-cabal2012].

@@

In this work, we have brought reaction norms to the genomic era by characterizing the thermal reactionomes of two temperate ant species, *Aphaenogaster picea* and *A. carolinensis*. At least 10% of their transcriptomes are thermally responsive. For both species, more genes responded to low than high temperatures. Together with physiological data showing more variation in critical minimum than maximum temperatures, characterization of the reactionome suggests that these species, and likely other ectotherms, have less adaptive potential in response to high than low temperatures. Our results also indicate that these two ant species have different responses to thermal extremes. *A. picea* responds by increasing metabolism and activating heat shock proteins and other protective molecules, whereas *A. carolinensis* shuts down metabolism and must rely on other mechanisms for thermal tolerance. The thermal reactionomes of these two species provide key insights into the genetic basis of thermal tolerance, and a resource for the future study of ecological adaptation in ant species. Finally, the reactionome itself illustrates a new direction for characterizing acclimation and adaptation in a changing climate.
